---
title: "Exercise 21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning = F)
```


## Combining the data
The data is available from Gelmans [website](http://stat.columbia.edu/~gelman/book/data/). The data combines multiple surveys and asks various questions. There is thus a whole range of columns in the data frame.
```{r, message=F, warning=F}
library(foreign)
library(tidyverse)
library(scales)

pre <- read.dta("pew_research_center_june_elect_wknd_data.dta")
election <- read.csv("2008ElectionResult.csv") %>%
  mutate(state=tolower(state))

colnames(pre)
```

We will concentrate here on the answer to question which political ideology the participant has:
```{r}
table(pre$ideo) / nrow(pre)
```

We want to estimate the percentage of the population in each state who label themselves as _very liberal_.
We also need the abbreviations for each state:
```{r}
mapping <- tibble(state = tolower(state.name)) %>%
   bind_cols(tibble(abb = state.abb)) %>% 
   bind_rows(tibble(state = c("washington dc", "district of columbia"), 
                    abb = c("DC", "DC")))

election <- election %>%
  left_join(mapping, by="state") %>%
  select(-state)
```

We aggregate the survey data by state so that we get a proportion estimate of liberals based on the survey:
```{r}
d <- pre %>% 
  filter( !is.na( ideo ) ) %>%
  select(state, ideo) %>%
  group_by(state) %>%
  summarise(lib=sum(ideo == "very liberal"), n=n()) %>%
  left_join(mapping, by="state") %>%
  left_join(election, by="abb") %>%
  mutate(lib.prop = lib / n ,
         obama_vote_share = vote_Obama_pct / 100,
         obama_won = vote_Obama_pct > 50)

d  
```

## Graph proportions vs vote-share
```{r, fig.height=6, fig.width=6}
d %>%
  ggplot(aes(x=lib.prop, y=obama_vote_share, label=abb, col=obama_won)) +
  geom_hline(yintercept = 0.5, col="grey", size=1) +
  geom_text(show.legend = F, size=3) +
  theme_minimal() +
  scale_color_brewer(palette="Set1") + 
  scale_y_continuous(labels = percent,
                     name = "Proportion of Obama Votes",
                     limits = c(0,1)) +
  scale_x_continuous(labels = percent,
                     name = "Proportion of Liberals in Survey",
                     limits = c(0, 0.15) )
```

We see that some states had a high proportion of liberals in the survey but a low proportion of votes for Obama in the election.

## Graph Bayes posterior mean vs vote-share
We assume that the number of liberals in the survey follows a Poisson distribution:
$$\text{Liberals}_j \sim \text{Poisson}(n_j\theta_j))$$ where $n_j$ number of Survey participants and $\theta$ is the underlying rate of liberals (that is, close to the proportion of votes for Obama) in state $j$.

The conjugate prior for the Poisson distribution is the Gamma distribution $\text{Gamma}(\alpha, \beta)$ with shape parameter $\alpha$ and rate parameter $\beta$.
We can use the Method of Moments to obtain estimates for our prior:
```{r}
( rate <- mean( d$lib / d$n ) / var( d$lib / d$n ) )
( shape <- rate*mean( d$lib / d$n ) )
```

This gives a prior that puts most probability mass between 0 and 0.1 for the rate parameter $\theta$.
```{r}
curve(dgamma(x, shape=shape, rate=rate), from=0, to=0.3)
hist(d$lib / d$n , add=T)
```

The posterior for Gamma is then $\text{Gamma}(\alpha + \sum_i^n y_i, \beta + \sum_i^n x_i)$ from which we can estimate the posterior mean.
```{r, fig.height=7, fig.width=7}
d %>%
  mutate(post_shape = shape + lib,
         post_rate = rate + n,
         post_mean = post_shape / post_rate ) %>%
  ggplot(aes(y=obama_vote_share, x=post_mean, label=abb, col=obama_won)) +
  geom_hline(yintercept = 0.5, col="grey", size=1) +
  geom_text(show.legend = F, size=3) +
  theme_minimal() +
  scale_color_brewer(palette="Set1") + 
  scale_y_continuous(labels = percent,
                     name = "Proportion of Obama Votes") +
  scale_x_continuous(labels = percent,
                     name = "Bayesian Mean of Proportion in Survey",
                     limits = c(0, 0.15) )
```




## Graph proportion vs number of survey-participants
```{r}
d %>%
  ggplot(aes(x=n, y=lib.prop, label=abb, col=obama_won)) +
  geom_text(show.legend = F, size=3) +
  theme_minimal() +
  scale_color_brewer(palette="Set1") +
  scale_x_continuous(name = "Number of Survey Participants") +
  scale_y_continuous(labels = percent,
                     name ="Proportion of Liberals in Survey",
                     limits = c(0,0.15))
```



## Graph Bayesian estimate vs number of survey-participants

```{r, fig.height=7, fig.width=7}
d %>%
  mutate(post_shape = shape + lib,
         post_rate = rate + n,
         post_mean = post_shape / post_rate ) %>%
  ggplot(aes(x=n, y=post_mean, label=abb, col=obama_won)) +
  geom_text(show.legend = F, size=3) +
  theme_minimal() +
  scale_color_brewer(palette="Set1") + 
  scale_x_continuous(name = "Number of Survey Participants") +
  scale_y_continuous(labels = percent,
                     name = "Bayesian Mean of Proportion in Survey",
                     limits = c(0, 0.15) )
```

