---
title: "r markdown"
author: "Salma Bouzid"
date: "4/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prior to solving this problem, section 1.6 of the book should be read. 

##4.a.Let's compute based on observed frequency:  
#### Pr(favorite wins | point spread = 8)
```{r cars}
number_fav_win = 8
total_num_games_played = 12

prob_fav_win_spr8 = number_fav_win/total_num_games_played
prob_fav_win_spr8
```

#### Pr(favorite wins by at least 8 | point spread = 8)
```{r cars}
number_fav_win_atleast_8 = 5
total_num_games_played = 12

prob_fav_win_atleast_8 = number_fav_win_atleast_8/total_num_games_played
prob_fav_win_atleast_8
```

#### Pr(favorite wins by at least 8 | point spread = 8 and favorite wins).

```{r cars}
number_fav_win_atleast_8 = 5
number_fav_win = 8

prob_fav_win_atleast_8_favorite_wins = number_fav_win_atleast_8/number_fav_win
prob_fav_win_atleast_8_favorite_wins
```

##4.(b) Let's build a paramteric model for the difference beween outcome and spread. 
In here we will assume that the distribution of the difference (outcome-spread) follows a normal distribution with mean 0 and standard deviation equal to 14. 

### Pr(favorite wins | point spread = 8)
### If our favorite team is to win, then he needs to have a point spread difference of more than 8, thus the 8.5

```{r cars}

prob_fav_win_spr8 <- pnorm(8.5,0,13.86)
prob_fav_win_spr8
```
#### Pr(favorite wins by at least 8 | point spread = 8)
```{r}
prob_fav_win_atleast_8 = pnorm(.5,0,14)
prob_fav_win_atleast_8
```

#### Pr(favorite wins by at least 8 | point spread = 8 and favorite wins) = P(favorite wins bz at least 8 | point spread = 8) / P(favorite wins by at least 8 | favorite wins)
```{r}
prob_fav_win_atleast_8/prob_fav_win_spr8
```

